---
title: "MockExam2"
author: "Erica Wurster"
date: "2025-03-06"
output: html_document
---

```{r setup, include=FALSE}
library(fpp2)
```

## Simple Moving Averages
```{r}
# Load necessary libraries
library(fpp2)

# Import the data
ts <- arrivals[,4]

# Plot the time series
plot(ts, main="International Arrivals to Australia from the US", xlab="Time (Quarterly)", ylab="Arrivals (thousands)")

# Compute moving averages
ma3 <- ma(ts, order=3)
ma6 <- ma(ts, order=6)
ma9 <- ma(ts, order=9)

# Add moving averages to the plot
lines(ma3, col="red", lwd=2)
lines(ma6, col="blue", lwd=2)
lines(ma9, col="green", lwd=2)
# As the moving average order increases, the forecast becomes smoother less impacted by the sharp peaks and valleys of the forecast. A higher order will be impacted less by the seasonality of the data.

# Bonus: Forecast for the next 12 months. MA3 is used as to not smooth out seasonality as best as possible. 
forecast_ma <- forecast(ma3, h=12)
plot(forecast_ma)
```

## Simple Smoothing
```{r}
# Perform simple exponential smoothing
ses_model <- ses(ts, h=12)
plot(ses_model)

# Extract model parameters
# Alpha: Measures how much weight is given to recent observations. About 35.8% of the forecasting weight is givn to the most recent observation.
ses_model$model$par["alpha"]
# Initial State: The estimated starting value of the time series.
ses_model$model$states[1, ]
# Sigma: Measure's the model's error variance; lower values indicate a better fit. The sigma level is very high, suggesting this model is not a good fit.
ses_model$model$sigma2

# Residual Analysis
residuals_ses <- residuals(ses_model)
# Plot of residuals: This indcates that the residuals constantly fluctuate around 0, there is no real pattern.
plot(residuals_ses)
# Histogram of Residuals: A greater amount of residuals are negative, the model tends to under predict rather than over predict.
hist(residuals_ses)
# Plot of fitted values vs residuals: The values are scattered around 0 with no clear pattern 
plot(as.numeric(fitted(ses_model)), residuals_ses)
# Plot of actuals vs residuals: Displays a positive correlation, suggesting that as the time series value increases, the residual also increases. The model may be consistently overpredicting the highest seasons.
plot(as.numeric(ts), residuals_ses)
# ACF of residuals: A high ACF occurs every other season, so the seasons with high visitors are driving the forecast up.
acf(na.omit(residuals_ses))

# Accuracy Measures
accuracy(ses_model)

# Forecast for Next Year
ses_forecast <- forecast(ses_model, h=4)
ses_forecast
plot(ses_forecast)
```
Summarize the forecasting technique:The high sigma value indicates that this model does not fit the forecast well. It predicts with 95% confidence that the forecast in one year will be between 129.0223 and 86.66809 thousand, with the point forecast being 112.2773 thousand. It tends to overpredict the seasons with lower visitors, a more seasonally adjusted forecast may be more suitable.
## Holt-Winters
```{r}
# Perform Holt-Winters forecasting
hw_model <- hw(ts, seasonal="additive", h=12)
plot(hw_model)

# Observations
alpha_hw <- hw_model$model$par["alpha"]
beta_hw <- hw_model$model$par["beta"]
gamma_hw <- hw_model$model$par["gamma"]

# Residual Analysis
residuals_hw <- residuals(hw_model)
plot(residuals_hw)
hist(residuals_hw)
plot(fitted(hw_model), residuals_hw)
plot(ts, residuals_hw)
acf(na.omit(residuals_hw))

# Accuracy Measures
accuracy(hw_model)

# Forecast for Next Year
hw_forecast <- forecast(hw_model, h=4)
hw_forecast
plot(hw_forecast)
```

## Accuracy Summary
```{r}
accuracy_table <- rbind(
  accuracy(ses_model),
  accuracy(hw_model)
)
rownames(accuracy_table) <- c("Simple Smoothing", "Holt-Winters")
accuracy_table
```

## Conclusion
Based on the forecasts, international arrivals are expected to increase slightly over the next year.

Over the next two years, a seasonal pattern suggests fluctuations with an upward trend.

Holt-Winters performed best in capturing seasonality, making it the preferred method for this dataset.

Simple exponential smoothing is useful for short-term trends but does not capture seasonality well.